Poisson approximate likelihood compared to the particle filter

### 2024-09-17 ms.Rnw

Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series \citep[Figure~3 of][]{hao24}.
By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrium distribution.
Line 5 of Table~\ref{tab:ovovrealdata} incorporates the SBH specification of initial values into the model of WWR.
The likelihood for this modified model was then maximized using an iterated filtering procedure similar to SBH.
Comparison of lines 3 and 5 shows that this improvement enables PF to reach the AIC values attained by PAL, and show a small improvement.
Further, at the parameter values obtained by maximizing with PF, we found that PF beats PAL (comparing lines 5 and 6).

>>>>

Inspection of log-likelihood anomalies \citep{wheeler24,li24} showed that the initial conditions for the latent process in January 2001 were fixed at values which were incompatible with the trajectory of the data early in the time series \citep[Figure~3 of][]{hao24}.
By contrast, SBH fixed their initial conditions 6 years before the first measurement, giving time for the system to reach its equilibrium distribution.
Line 5 of Table~\ref{tab:ovovrealdata} incorporates the SBH specification of initial values into the model of WWR.
The likelihood for this modified model was then maximized using an iterated filtering procedure similar to SBH.
Comparison of lines 3 and 5 shows that this improvement enables PF to reach the AIC values attained by PAL, and show a small improvement.
Further, at these maximum likelihood parameter values, we found that PF beats PAL (comparing lines 5 and 6; for line 6 we initialized PAL at the average value of the stochastic initialization used in line 5).
